Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2025-06-09 paper.tex
Second round of revisions

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves
significantly influences remote sensing, interpretation of \textit{in situ} wave
measurements, and numerical wave forecasting.
This paper revisits the wave crest and action conservation laws and derives
steady, nonlinear, analytical solutions for the change of short-wave wavenumber,
action, and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full 
conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed 5/8 to
wavenumber, 1/4 due to wave action, and 1/8 due to effective gravity.
Examining the homogeneity and stationarity requirements for the conservation of
wave action reveals that stationarity is a stronger requirement and is
generally not satisfied for very steep long waves.
We examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and not generally.
The proposed steady solutions are a good approximation of the nonlinear
crest-action conservation solutions in long-wave steepness $\lesssim 0.2$.
Except for a subset of initial conditions, the solutions to the non-linearised
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:dispersion}
\omega = \sqrt{gk} + k U
\end{equation}
where $\omega$ is the angular frequency in a fixed reference frame, $g$ is the
gravitational acceleration, $k$ is the wavenumber, and $U$ is the mean advective
current in the direction of the wave propagation.
A current in the direction of the waves increases their absolute frequency
without changing their wavenumber.
An important limitation here is that $U$ must be slowly varying on the scales of
the wave period \citep{bretherton1968wavetrains}.
From the perspective of short waves riding on long waves, the advective current
is the horizontal near-surface orbital velocity of the long wave.
Strictly, a velocity profile must be considered when determining the
advective velocity for a wave of any given wavenumber \citep{stewart1974hf},
however, for brevity, here we will assume a surface velocity to be the advective
one for short waves.
The evolution of the short-wave wavenumber is described by the conservation of
wave crests \citep[e.g.,][]{whitham1974dispersive,phillips1981dispersion}:

@@

\begin{equation}
\label{eq:wave_action}
\dfrac{\partial N}{\partial t}
+ \dfrac{\partial}{\partial x} \left[\left(C_g + U\right)N\right]
= 0
\end{equation}
where $N$ is the action of short waves defined as the ratio of their
energy to their intrinsic frequency, $E/\sigma$,
and $C_g = 1/2\sqrt{g/k}$ is their group speed in deep water.
The wave action expressed as $E/\sigma$ is commonly taken as
conservative in wave modulation theory
\citep{phillips1981dispersion,longuet1987propagation}, however as
\citet{dysthe1988orbiting} show, it is conservative only for gently sloped waves.
We describe this and other requirements in more detail further below and show
in what long-wave conditions they are satisfied.
The wave crest conservation (\ref{eq:wave_crests}) follows from the assumption
of a conserved wave phase, which is limited by the same set of requirements as
those for the wave action conservation \citep[e.g.,][]{whitham1974dispersive}.
Although the absence of wave growth and dissipation is not generally realistic in
the open ocean, it is a useful approximation here to isolate the effects
of hydrodynamic modulation solely due to the motion of the long waves.

@@

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}
This equation is similar to (\ref{eq:wavenumber}), except for the last term
which represents the inhomogeneity of the group speed.
As \citet{peureux2021unsteady} explained, the presence of this term causes
unsteady growth of wave action in infinite long-wave trains.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are the governing equations
that we approximate and solve analytically in \S\ref{section:analytical_solutions},
and numerically in their full form in \S\ref{section:numerical_solutions}.
This system of equations is semi-coupled, meaning that
it is possible to solve (\ref{eq:wavenumber}) on its own, but solving
(\ref{eq:wave_action2}) requires also the solution of (\ref{eq:wavenumber}).
In other words, the wave kinematics (the wavenumber distribution) are not
concerned with the wave action.
In contrast, the wave action is governed, among others, by the
group velocity and thus depends on the wavenumber.

@@

\begin{equation}
  \sigma_L \ll \widetilde{\sigma}
\end{equation}
\citet{bretherton1968wavetrains} also require that the medium is nearly uniform
over the vertical scales of motion of the short waves.
For the ambient velocity being the orbital velocity of the long wave, this is
trivially satisfied with (\ref{eq:k_scale_separation}).

@@

Fig. \ref{fig:homogeneity_stationarity_analytical} shows the homogeneities and
stationarities of the short-wave wavenumber and effective gravity based on
the steady solutions and their derived criteria as minima of
(\ref{eq:homogeneity_condition})-(\ref{eq:stationarity_condition}).
Stationarity is a stronger requirement than homogeneity.
At the lowest scale separation considered ($k/k_L = 10$), short-wave wavenumber,
action, and effective gravity are all strongly homogeneous ($H_{k,N,g} > 0.99$)
for $\varepsilon_L < 0.1$.
However, the solutions are only strongly stationary ($S_{k,N,g} > 0.99$) at a
high scale separation of $k/k_L = 100$ for $\varepsilon_L < 0.1$.
If we consider weakly stationary conditions as $S_{k,N,g} > 0.9$, then the
solutions satisfy the requirements for the wave action conservation for
$\varepsilon_L < 0.3$ at $k/k_L = 10$, and for $\varepsilon_L < 0.4$ at
$k/k_L > 20$.
Although \citet{bretherton1968wavetrains} state that the wave action balance
is valid to $\mathcal{O}(k_L/k)$, applying the general criteria
(\ref{eq:homogeneity_condition})-(\ref{eq:stationarity_condition}) yields an
error of $\mathcal{O}(\varepsilon_L k_L/k)$.
We will revisit these criteria based on the numerical solutions of nonlinear
wave action and crest conservation equations in the next section.

@@

In this paper we revisited the problem of short-wave modulation by long waves
from a wave action conservation perspective.
Using the linearised wave crest and action conservation equations coupled with
nonlinear solutions for the effective gravity, we derived steady, nonlinear, analytical
solutions for the modulation of short wave wavenumber, amplitude, steepness,
intrinsic frequency, and phase speed.
The latter two were previously examined in the context of short-wave
modulation \citep{longuet1962resonant,longuet1962phase}, and may be important
for correctly calculating the wind input into short waves, which is proportional
to the wind speed relative to the wave celerity.
The approximate steady solutions yield higher wavenumber and amplitude
modulation than that of \citet{longuet1960changes}, and somewhat lower than 
the numerical solutions of \citet{longuet1987propagation} and
\citet{zhang1990evolution}.
This is mainly due to the analytical solution requiring the crest and action
conservation equations to be linearised, and the nonlinearity arising due
to the evaluation of short waves on the surface of the long wave (as opposed to
the mean water level) and due to nonlinear effective gravity.
In the steady solutions, the short-wave steepness modulation is dominated by
the wavenumber modulation (5/8), followed by the wave action modulation (1/4),
and is least affected by the effective gravity modulation (1/8).
The homogeneity and stationarity criteria for wave action conservation are
evaluated for both the analytical and numerical solutions.
The stationarity is a stronger criterion than homogeneity, and is found to be
strongly satisfied ($> 99\%$ stationary) for very gently sloped waves and weakly
satisfied ($> 90\%$ stationary) for steeper waves.

@@

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves
significantly influences remote sensing, interpretation of \textit{in situ} wave
measurements, and numerical wave forecasting.
This paper revisits the wave crest and action conservation laws and derives
steady, nonlinear, analytical solutions for the change of short-wave wavenumber,
action, and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical solutions of the full 
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed 5/8 to
wavenumber, 1/4 due to wave action, and 1/8 due to effective gravity.
Examining the homogeneity and stationarity requirements for the conservation of
wave action reveals that stationarity is a stronger requirement and is
generally not satisfied for very steep long waves.
We examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and not generally.
The proposed steady solutions are a good approximation of the nonlinear
crest-action conservation solutions in long-wave steepness $\lesssim 0.2$.
Except for a subset of initial conditions, the solutions to the non-linearised
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:dispersion}
\omega = \sqrt{gk} + k U
\end{equation}
where $\omega$ is the angular frequency in a fixed reference frame, $g$ is the
gravitational acceleration, $k$ is the wavenumber, and $U$ is the mean advective
current in the direction of the wave propagation.
A current in the direction of the waves increases their absolute frequency
without changing their wavenumber.
An important limitation here is that $U$ must be slowly varying on the scales of
the wave period \citep{bretherton1968wavetrains}.
From the perspective of short waves riding on long waves, the advective current
is the horizontal near-surface orbital velocity of the long wave.
Strictly, shear must be considered when determining the advective velocity for
a wave of any given wavenumber \citep{stewart1974hf,ellingsen2014ship}.
Here, we assume the surface velocity to be the advective one for short waves,
and demonstrate later in $\S$\ref{sec:wave_action_conservation} that this is
valid for sufficient scale separation between the short and long waves.
The evolution of the short-wave wavenumber is described by the conservation of
wave crests \citep[e.g.,][]{whitham1974dispersive,phillips1981dispersion}:

@@

\begin{equation}
\label{eq:wave_action}
\dfrac{\partial N}{\partial t}
+ \dfrac{\partial}{\partial x} \left[\left(C_g + U\right)N\right]
= 0
\end{equation}
where $N$ is the action of short waves defined as the ratio of their
energy to their intrinsic frequency, $E/\sigma$,
and $C_g = 1/2\sqrt{g/k}$ is their group speed in deep water.
The wave action expressed as $E/\sigma$ is commonly taken as
conservative in wave modulation theory
\citep{phillips1981dispersion,longuet1987propagation}, however as
\citet{dysthe1988orbiting} show, it is conservative only for gently sloped waves.
We describe this and other requirements in more detail further below and show
in what long-wave conditions they are satisfied.
The wave crest conservation (\ref{eq:wave_crests}) follows from the assumption
of a conserved wave phase, and is another prerequisite for wave action
conservation \citep[e.g.,][]{whitham1974dispersive}.
Although the absence of wave growth and dissipation is not generally realistic in
the open ocean, it is a useful approximation here to isolate the effects
of hydrodynamic modulation solely due to the motion of the long waves.
The governing equations (\ref{eq:dispersion})-(\ref{eq:wave_action}) are
derived in Appendix \ref{appendix:derivation}.

@@

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}
This equation is similar to (\ref{eq:wavenumber}), except for the last term
which represents the inhomogeneity of the group speed.
As \citet{peureux2021unsteady} explained, the presence of this term causes
unsteady growth of wave action in infinite long-wave trains.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are the governing equations
that we approximate and solve analytically in \S\ref{section:analytical_solutions},
and numerically in their full form in \S\ref{section:numerical_solutions}.
This system of equations is semi-coupled, meaning that
it is possible to solve (\ref{eq:wavenumber}) on its own, but solving
(\ref{eq:wave_action2}) requires also the solution of (\ref{eq:wavenumber}).
In other words, the wave kinematics (the wavenumber distribution) are not
concerned with the wave action.
In contrast, the wave action is governed, among others, by the
group velocity and thus depends on the wavenumber.
This convenient property is only possible if we consider the linear form
of the dispersion relation, that is, if the series are truncated at the
second-order solution in the Stokes expansion.
For a third-order expansion, an additional term that is proportional to the
square of the wave steepness appears in the dispersion relation.
In that case, equations (\ref{eq:wavenumber}) and (\ref{eq:wave_action2})
become fully coupled and the analytical solution for the wavenumber is no longer
feasible.

@@

\begin{equation}
  \sigma_L \ll \widetilde{\sigma}
\end{equation}
\citet{bretherton1968wavetrains} also require that the medium is nearly uniform
over the vertical scales of motion of the short waves:

\begin{equation}
  \frac{1}{U} \frac{\partial U}{\partial z} \ll \widetilde{k}
\end{equation}
which, for a long deep water wave, is equivalent to (\ref{eq:k_scale_separation}).
Alternatively, if we use the second Froude number criterion following
\citet{ellingsen2014ship} and quantify the shear length
under the long wave surface as:

\begin{equation}
  l_S = \frac{g}{S^2} \approx \frac{g}{\varepsilon_L^2 \omega_L^2}
\end{equation}
we arrive to an additional requirement:

\begin{equation}
  \frac{k}{k_L} \gg \varepsilon_L^2
\end{equation}
which relates the scale separation to the long-wave steepness.

@@

Fig. \ref{fig:homogeneity_stationarity_analytical} shows the homogeneities and
stationarities of the short-wave wavenumber and effective gravity based on
the steady solutions and their derived criteria as minima of
(\ref{eq:homogeneity_condition})-(\ref{eq:stationarity_condition}).
Stationarity is a stronger requirement than homogeneity.
At the lowest scale separation considered ($k/k_L = 10$), short-wave wavenumber,
action, and effective gravity are all strongly homogeneous ($H_{k,N,g} > 0.99$)
for $\varepsilon_L < 0.1$.
However, the solutions are only strongly stationary ($S_{k,N,g} > 0.99$) at a
high scale separation of $k/k_L = 100$ for $\varepsilon_L < 0.1$.
If we consider weakly stationary conditions as $S_{k,N,g} > 0.9$, then the
solutions satisfy the requirements for the wave action conservation for
$\varepsilon_L < 0.3$ at $k/k_L = 10$, and for $\varepsilon_L < 0.4$ at
$k/k_L > 20$.
Although \citet{bretherton1968wavetrains} state that the wave action balance
is valid to $\mathcal{O}(k_L/k)$, applying the general criteria
(\ref{eq:homogeneity_condition})-(\ref{eq:stationarity_condition}) yields an
error of $\mathcal{O}(\varepsilon_L k_L/k)$.
We will revisit these criteria based on the numerical solutions of nonlinear
wave action and crest conservation equations in the next section.
For the asymptotic limits of wave action conservation in terms of the wave
short-wave steepness, see Appendix \ref{appendix:wave_action_conservation}.

@@

In this paper we revisited the problem of short-wave modulation by long waves
from a wave action conservation perspective.
Using the linearised wave crest and action conservation equations coupled with
nonlinear solutions for the effective gravity, we derived steady, nonlinear, analytical
solutions for the modulation of short wave wavenumber, amplitude, steepness,
intrinsic frequency, and phase speed.
The latter two were previously examined in the context of resonant interactions
\citep{longuet1962resonant,longuet1962phase}, a modulation process distinct from
the one discussed here.
Both effects, however, may be important for correctly calculating the wind input
into short waves, which is proportional to the wind speed relative to the wave
celerity.
The approximate steady solutions yield higher wavenumber and amplitude
modulation than that of \citet{longuet1960changes}, and somewhat lower than
the numerical solutions of \citet{longuet1987propagation} and
\citet{zhang1990evolution}.
This is mainly due to the analytical solution requiring the crest and action
conservation equations to be linearised, and the nonlinearity arising due
to the evaluation of short waves on the surface of the long wave (as opposed to
the mean water level) and due to nonlinear effective gravity.
In the steady solutions, the short-wave steepness modulation is dominated by
the wavenumber modulation (5/8), followed by the wave action modulation (1/4),
and is least affected by the effective gravity modulation (1/8).
The homogeneity and stationarity criteria for wave action conservation are
evaluated for both the analytical and numerical solutions.
The stationarity is a stronger criterion than homogeneity, and is found to be
strongly satisfied ($> 99\%$ stationary) for very gently sloped waves and weakly
satisfied ($> 90\%$ stationary) for steeper waves.

@@

\section{Derivation of the governing equations}
\label{appendix:derivation}

Here we derive the governing equations for the short waves
(\ref{eq:dispersion})-(\ref{eq:wave_action}) and discuss their asymptotic
consistency.

\subsection{Wave dispersion}

An irrotational flow has a velocity potential such that the continuity can be
expressed as:

\begin{equation}
  \nabla^2 \phi = 0
  \label{eq:laplace_equation}
\end{equation}
Boundary conditions that allow solving for $\phi$ are the dynamic and kinematic
free surface boundary conditions, respectively:

\begin{equation}
  \frac{\partial \phi}{\partial t} + \frac{1}{2} (\nabla{\phi})^2 + gz = 0 \quad \text{at} \quad z = \eta(x,t)
  \label{eq:dynamic_free_surface_boundary_condition}
\end{equation}

\begin{equation}
  \frac{\partial \eta}{\partial t} + \frac{\partial \phi}{\partial x} \frac{\partial \eta}{\partial x} + \frac{\partial \phi}{\partial z} = 0 \quad \text{at} \quad z = \eta(x,t)
  \label{eq:kinematic_free_surface_boundary_condition}
\end{equation}
Also required is the bottom boundary condition:

\begin{equation}
  \frac{\partial \phi}{\partial z} = 0 \quad \text{at} \quad z = -h
\end{equation}
In the \citet{stokes1847} perturbation approach, the velocity potential and
elevation are expanded in a power series of the wave steepness
$\varepsilon = ak$ as:

\begin{equation}
  \phi = \sum_{n=1}^{\infty} \varepsilon^n \phi_n
\end{equation}
\begin{equation}
  \eta = \sum_{n=1}^{\infty} \varepsilon^n \eta_n
\end{equation}
The power series expansion requires $\varepsilon \ll 1$.
The velocity potential and elevation series are truncated at the desired order
and inserted into the dynamic free surface boundary condition (\ref{eq:dynamic_free_surface_boundary_condition}).
Up to the third order, the solutions for the velocity potential and the
corresponding velocity components are (\ref{eq:phi})-(\ref{eq:W_L}), and the
surface elevation is (\ref{eq:eta_stokes}).
Evaluating these quantities to higher orders in $\varepsilon$ is possible at
the expense of higher complexity and with diminishing returns in accuracy.

Combining the solutions for the potential and the surface elevation with the
kinematic free surface boundary condition (\ref{eq:kinematic_free_surface_boundary_condition})
yields the dispersion relation, which in the linear case and in deep water is
(\ref{eq:dispersion}).
At the third order in $\varepsilon$, the frequency begins to depend on the wave
steepness $\varepsilon$ as well:

\begin{equation}
  \omega = \sqrt{g k} \left(1 + \frac{1}{2} \varepsilon^2 + \mathcal{O}(\varepsilon^4)\right) + U k
  \label{eq:dispersion_nonlinear}
\end{equation}
In this paper we constrain the analysis framework to the linear case for short
waves, as this permits the analytical solutions in $\S\ref{section:analytical_solutions}$.
The Doppler shift by the ambient velocity $U$ is unaffected by the nonlinear
effects in the solution.

\subsection{Conservation of wave crests}

A system with a well defined phase function $\psi(x,t)$ also has the wavenumber
and frequency defined as its gradients with respect to space and time, respectively:

\begin{equation}
  k = \frac{\partial \psi}{\partial x}
  \label{eq:wavenumber_from_phase}
\end{equation}

\begin{equation}
  \omega = - \frac{\partial \psi}{\partial t}
  \label{eq:frequency_from_phase}
\end{equation}
implying $\psi = kx - \omega t$.
Differentiating (\ref{eq:wavenumber_from_phase})-(\ref{eq:frequency_from_phase})
with respect to time and space, respectively, and then eliminating $\psi$ yields:

\begin{equation}
  \frac{\partial k}{\partial t} + \frac{\partial \omega}{\partial x} = 0
  \label{eq:wavenumber_conservation}
\end{equation}
which is the crest conservation equation (\ref{eq:wave_crests}).
That this is the wavenumber conservation equation with the group speed as the
transport velocity can be recognized by writing:

\begin{equation}
  \frac{\partial k}{\partial t} + \frac{\partial \omega}{\partial k} \frac{\partial k}{\partial x} = 0
\end{equation}
This conservation equation is valid regardless of the order of nonlinearity
considered, as long as $\psi$ is well defined.
Nonlinearity in terms of $\varepsilon$ can, however, enter (\ref{eq:wavenumber_conservation})
through nonlinear forms of $\omega$ such as (\ref{eq:dispersion_nonlinear}).

\subsection{Conservation of wave action}
\label{appendix:wave_action_conservation}

The conservation of wave action (\ref{eq:wave_action}) is most readily derived
following the variational approach by \citet{whitham1965general}.
Begin with a statement that infinitesimal variations of a Lagrangian of a system
over a spatial domain and time interval must vanish:

\begin{equation}
  \delta \iint L\, dx\,dt = 0
\end{equation}
Here, $L$ is the instantaneous Lagrangian density, a scalar quantity that
corresponds to the difference between the kinetic and potential energy of the
system:

\begin{equation}
  L = \int_{-h}^{\eta} \left[ \frac{1}{2} (\nabla{\phi})^2 + gz \right] dz
  \label{eq:lagrangian_density_energy_difference}
\end{equation}
\citet{luke1967variational} showed that the equations for irrotational
surface waves follow from the variational approach if the Lagrangian $L$ is
defined instead as:

\begin{equation}
  L = \int_{-h}^{\eta} \left[ \frac{\partial \phi}{\partial t} + \frac{1}{2} (\nabla{\phi})^2 + gz \right] dz
  \label{eq:lagrangian_density_luke}
\end{equation}
(\ref{eq:lagrangian_density_energy_difference}) and (\ref{eq:lagrangian_density_luke})
are equivalent in the sense that they both yield (\ref{eq:laplace_equation}),
however, varying the latter is necessary to attain the correct boundary conditions
\citep{whitham1967non}.
Notice that for water waves the instantaneous Lagrangian being zero is
equivalent to the Bernoulli's equation, which is also the dynamic free surface
boundary condition.

Then, Whitham's averaged variation principle is obtained by averaging the
instantaneous Lagrangian $L$ over one wave period:

\begin{equation}
  \mathcal{L} = \frac{1}{2\pi} \int_0^{2\pi} L\ d\psi
  \label{eq:averaged_lagrangian}
\end{equation}
where $\mathcal{L}$ is the averaged Lagrangian density.
The averaged variational principle then states:

\begin{equation}
  \delta \iint \mathcal{L}\, dx\,dt = 0
\end{equation}
The averaged Lagrangian for water waves is obtained by integrating
(\ref{eq:lagrangian_density_luke}) over depth, and then averaging over the phase
(\ref{eq:averaged_lagrangian}).
In the deep water limit, it is:

\begin{equation}
  \mathcal{L} = \frac{1}{4} \left( 1 - \frac{\omega^2}{gk} \right) g a^2 + \mathcal{O}(\varepsilon^4)
\end{equation}
which, after recognizing the wave energy$E = \frac{1}{2} g a^2$, yields:

\begin{equation}
  \mathcal{L} = \frac{1}{2} \left( 1 - \frac{\omega^2}{gk} \right) E + \mathcal{O}(\varepsilon^4)
  \label{eq:averaged_lagrangian_deep_water}
\end{equation}

Whitham's conservation equations for water waves are:

\begin{equation}
  \frac{\partial \mathcal{L}}{\partial a} = 0
\end{equation}

\begin{equation}
  \frac{\partial}{\partial t} \left(\frac{\partial \mathcal{L}}{\partial \omega}\right) - \frac{\partial}{\partial x} \left(\frac{\partial \mathcal{L}}{\partial k}\right) = 0
  \label{eq:whitham_conservation_equation}
\end{equation}
Inserting (\ref{eq:averaged_lagrangian_deep_water}) into (\ref{eq:whitham_conservation_equation})
and defining:

\begin{equation}
  \frac{\partial \mathcal{L}}{\partial \omega} = \frac{E}{\sigma} \equiv N
  \label{eq:adiabatic1}
\end{equation}
yields the conservation of wave action:

\begin{equation}
  \frac{\partial}{\partial t} \left[ N + \mathcal{O}(\varepsilon^4) \right] +
  \frac{\partial}{\partial x} \left[ (C_g + \mathcal{O}(\varepsilon^2)) N + \mathcal{O}(\varepsilon^4) \right] = 0
\end{equation}
Due to the $\mathcal{O}(\varepsilon^2)$ correction to the group speed, the
overall truncation error for the wave action balance is $\mathcal{O}(\varepsilon^2)$
as well:

\begin{equation}
  \frac{\partial N}{\partial t} +
  \frac{\partial (C_g N)}{\partial x} = \mathcal{O}(\varepsilon^2)
  \label{eq:wave_action_conservation_nonlinear}
\end{equation}
which was introduced as a governing equation in \S\ref{section:governing_equations}.
Although the truncation error of $\mathcal{O}(\varepsilon^2)$ arising from $C_g$
seems to have been omitted in the original derivation by \citet{whitham1967non},
it appeared correctly later in \citet{whitham1974linear}.
An important nuance about applying the nonlinear dispersion here is that the
secondary (wave-induced) mean flow and set-up effects can be ignored in the
deep water limit; this is discussed by \citet{whitham1974dispersive}.
The terms proportional to $\varepsilon^4$ do not appear in the averaged Lagrangian
(\ref{eq:averaged_lagrangian_deep_water}) until the third-order Stokes waves
are considered, for which the dispersion gains a $\mathcal{O}(\varepsilon^2)$
correction.
In that case, the conserved quantity is:

\begin{equation}
  \frac{\partial \mathcal{L}}{\partial \omega} = \frac{E}{\sigma} (1 + \frac{1}{2} \varepsilon^2)
  \label{eq:adiabatic2}
\end{equation}
and the group speed gains a nonlinear factor in $\varepsilon^2$ as well.
(\ref{eq:wave_action_conservation_nonlinear}), then, maintains the same form,
but with the truncation error of $\mathcal{O}(\varepsilon^4)$.
This derivation, as well as the averaged Lagrangian principle applied to the
wave-induced mean flow over arbitrary water depths, is detailed at greater length
by \citet{whitham1967non}.

To consider the effect of slowly varying ambient currents,
\citet{bretherton1968wavetrains} carried out the same derivation except that
they propagated the ambient current derivatives $\partial U / \partial t$ and
$\partial U / \partial x$ through the averaged Lagrangian.
To allow these derivatives to drop out of the averaging integral, the authors
make an argument that the variation must be sufficiently small, no larger than
$\mathcal{O}(\frac{1}{\sigma U} \frac{\partial U}{\partial t})$ and
$\mathcal{O}(\frac{1}{k U} \frac{\partial U}{\partial x})$, respectively.
Taking into consideration the asymptotic limits in terms of the wave steepness
$\varepsilon$ as well as the variation of the ambient current, the wave action
balance can be written as:

\begin{equation}
  \frac{\partial N}{\partial t} +
  \frac{\partial}{\partial x} \left[ \left(C_g + U\right) N \right] =
  \mathcal{O}\left(\varepsilon^4, \frac{1}{\sigma U} \frac{\partial U}{\partial t}, \frac{1}{k U} \frac{\partial U}{\partial x}\right)
\end{equation}
\S\ref{sec:wave_action_conservation} deals specifically with quantifying the
ambient velocity variation in the context of long-wave orbital velocities.

There is an alternative derivation of the wave action conservation that was
pointed out by \citet{bretherton1968wavetrains}.
Starting from the wave energy conservation in non-uniform currents derived by
\citet{longuet1961changes}

\begin{equation}
  \frac{\partial E}{\partial t} +
  \frac{\partial}{\partial x} \left[ \left(C_g + U\right) E \right] +
  S_{x} \frac{\partial U}{\partial x} = 0
  \label{eq:energy_conservation_nonuniform_currents}
\end{equation}
where $S_x$ is the wave radiation stress, and combining it with the dispersion
relation (\ref{eq:dispersion}) and conservation of crests
(\ref{eq:wave_crests}), they show that (\ref{eq:energy_conservation_nonuniform_currents})
is equivalent to the wave action conservation (\ref{eq:wave_action}).
This derivation is found in the Appendix of \citet{bretherton1968wavetrains},
and was later recognized by \citet{whitham1974linear} as well.

In summary, asymptotically consistent sets of dispersion and action conservation
equations vary depending on the level of nonlinearity considered.
In the case of linear waves, the dispersion relation (\ref{eq:dispersion}),
the averaged Lagrangian (\ref{eq:averaged_lagrangian_deep_water}), the adiabatic
quantity (\ref{eq:adiabatic1}) and the wave action conservation
(\ref{eq:wave_action_conservation_nonlinear}) are asymptotically consistent.
For third-order Stokes waves, the dispersion relation
(\ref{eq:dispersion_nonlinear}), the adiabatic quantity (\ref{eq:adiabatic2}),
together with the group speed that includes a $\mathcal{O}(\varepsilon^2)$ correction,
are governed by the action conservation law
(\ref{eq:wave_action_conservation_nonlinear}) and are asymptotically consistent.
This is summarized in Table \ref{table:asymptotic_consistency}.

\begin{table}
\begin{center}
\def~{\hphantom{0}}
\begin{tabular}{cccc}
Order & Frequency  & Group speed & Wave action conservation \\
\hline
1, 2 & $\sqrt{gk}(1 + \mathcal{O}(\varepsilon^2))$ & $\frac{1}{2}\sqrt{\frac{g}{k}}(1 + \mathcal{O}(\varepsilon^2))$ & $\frac{\partial N}{\partial t} + \frac{\partial (C_g N)}{\partial x} = \mathcal{O}(\varepsilon^2)$ \\
\hline
3 & $\sqrt{gk}(1 + \frac{1}{2}\varepsilon^2 + \mathcal{O}(\varepsilon^4))$ & $\frac{1}{2}\sqrt{\frac{g}{k}}(1 + \frac{5}{2}\varepsilon^2 + \mathcal{O}(\varepsilon^4))$ & $\frac{\partial N}{\partial t} + \frac{\partial (C_g N)}{\partial x} = \mathcal{O}(\varepsilon^4)$ \\
\hline
\end{tabular}
\caption{
  Asymptotically consistent sets of dispersion and wave action relations for
  deep water and their respective truncation errors in terms of the wave
  steepness $\varepsilon$, for the first three Stokes expansion orders.
}
\label{table:asymptotic_consistency}
\end{center}
\end{table}
